\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-27 paper.tex
Clarify arg-unary

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and an ARG that we can derive from a sequence of \emph{estimated}
local trees.
The ARG inference method
\espalier~\citep{rasmussen2022espalier} is illustrative in this context.
It begins by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees.
\espalier\ approaches this task
by identifying maximimal subtrees
that do not change between pairs of adjacent
local trees and then
heuristically exploring the space of possible
rearrangements of these subtrees.
To derive precise details about recombination events,
\espalier\ then attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
% A similar problem is tackled by \arbores~\citep{heine2018bridging}, which
% explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing
% and resimulating candidate SPR moves connecting two fixed, leaf-labelled local
% trees.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.

@@

This distinction between ARG and local arity is mainly
of interest when we consider those that have
a single child: that are \emph{unary}.
Returning to the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h}
are ARG-unary (Fig.~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig.~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{f} has a single child in this
interval: in other words it is a \emph{locally unary} node.
In another example, \noderef{e} is binary in the graph, being a common
ancestor of \noderef{a} and \noderef{c}, but is
locally unary in all trees in which it is present.
This is because no ancestral material coalesces there:
\noderef{a} inherits genetic material from the far left hand end of
\noderef{e},
while \noderef{c} only inherits the (disjoint) right hand end.

[FIXME this isn't true, add comment about eARG conversion]
Outside of the context of simulation, ARG-unary nodes are
only likely to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-CoV-2; see Discussion).
In contrast, locally unary nodes, are a common and important feature
of many different types of ARG.
As discussed in the previous section,
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them. As we see in the next section, both ARG and locally
unary nodes occur in a range of different scenarios.

>>>>

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and an ARG that we can derive from a sequence of \emph{estimated}
local trees.
The ARG inference method
\espalier~\citep{rasmussen2022espalier} is illustrative in this context.
It begins by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees.
\espalier\ approaches this task
by identifying maximimal subtrees
that do not change between pairs of adjacent
local trees and then
heuristically exploring the space of possible
rearrangements of these subtrees.
To derive precise details about recombination events,
\espalier\ then attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
% NOTE: dropping this because it doesn't clarify the actual point.
% A similar problem is tackled by \arbores~\citep{heine2018bridging}, which
% explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing
% and resimulating candidate SPR moves connecting two fixed, leaf-labelled local
% trees.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.

@@

This distinction between ARG and local arity is mainly
of interest when we consider nodes that have
a single child: those that are \emph{unary}.
Returning to the example in Fig.~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h}
are ARG-unary (Fig.~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig.~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, node \noderef{f} has a single child in this
interval: it is \emph{locally unary}.
In another example, \noderef{e} is binary in the graph, being a common
ancestor of \noderef{a} and \noderef{c}, but is
locally unary in all trees in which it is present.
This is because no ancestral material coalesces in \noderef{e}:
\noderef{a} inherits genetic material from the far left hand end of
\noderef{e},
while \noderef{c} only inherits the (disjoint) right hand end.

ARG-unary nodes occur when an ARG node has one child and one or
more parents.
Nodes with one child and one parent (``pass-through'' nodes)
occur where we wish the record the passage of ancestral material
through a known node. For example, in simulations
it is sometimes useful to record the passage of
ancestral material through known pedigree individuals
regardless of whether common ancestry occurs.
ARG-unary nodes are also occur in inferences
from longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-CoV-2; see Discussion).
Nodes with one child and two parents arise when
we model a recombination event using a single node
in the classical manner
(e.g.\ Fig.~\ref{fig-ancestry-resolution}).

Locally unary nodes are a common and important feature
of many different types of ARG.
As discussed in the previous section,
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them. As we see in the next two sections, both ARG and locally
unary nodes occur in various scenarios, and are
produced by a range of current inference methods.
